---
title: "p8105_hw2_jy3270"
author: "Jingya Yu"
date: "2022-10-04"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# load all libraries
library(tidyverse)
library(readr)
library(readxl)
library(tidyr)
```

# Problem 1
#### read in data
```{r}
original <- read_csv(file = "./NYC_Transit_Subway_Entrance_And_Exit_Data.csv") %>% janitor::clean_names()
```
- clean data

```{r}
nyc_transit <- original %>%
  select(line, station_name, station_latitude:entry, vending, ada) %>%
  mutate(entry = recode(entry, "YES" = TRUE, "NO" = FALSE))
head(nyc_transit)
```

The original dataset contains `r nrow(original)` rows and `r ncol(original)` columns. The variable names are: Division, Line, Station Name, Station Latitude, Station Longitude, Route1, Route2, Route3, Route4, Route5, Route6, Route7, Route8, Route9, Route10, Route11, Entrance Type, Entry, Exit Only, Vending, Staffing, Staff Hours, ADA, ADA Notes, Free Crossover, North South Street, East West Street, Corner, Entrance Latitude, Entrance Longitude,	Station Location and Entrance Location. Out of these variables, I retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance.

The cleaned dataset contains `r nrow(nyc_transit)` rows and `r ncol(nyc_transit)` columns. For data cleaning, I first clean the names of data using janitor package, then only select the required variables, and change the entry variable from character to logical variables. The data is not very clean because the route variables can be further reorganized. At the same time, some ofthe variable names contain double and have not been changed yet.

- number of distinct station

```{r}
nyc_transit %>%
  distinct(line, station_name) %>%
  nrow()
```

There are 465 distinct stations.

- number of ADA compliant stations

```{r}
nyc_transit %>%
  filter(ada == TRUE) %>%
  distinct(line, station_name) %>%
  nrow()
```

There are 84 ADA compliant stations.

- proportion of station entrances / exits without vending 

```{r}
nyc_transit %>%
  filter(vending == "NO") %>%
  pull(entry)%>%
  mean()
```

There are 37.7% of the stations entrances / exits without vending.

- number of stations serving A
```{r}
# first change all the route names to character
refored_nyc_transit = mutate(nyc_transit,across(route1:route11, as.character))

refored_nyc_transit %>%
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route") %>% 
  filter(route == "A") %>% 
  select(station_name, line) %>% 
  distinct %>% 
  nrow()
```
There are 60 stations serving A

- A train with ADA compliant
```{r}
refored_nyc_transit %>% 
  pivot_longer(route1:route11, names_to = "route_num", values_to = "route") %>%
  filter(route == "A", ada == TRUE) %>% 
  select(station_name, line) %>% 
  distinct() %>%
  nrow()
```

There are 17 stations serving A with ADA compliant.

### Problem 1(answer key attached)
Below we import and clean data from `NYC_Transit_Subway_Entrance_And_Exit_Data.csv`. The process begins with data import, updates variable names, and selects the columns that will be used in later parts fo this problem. We update `entry` from `yes` / `no` to a logical variable. As part of data import, we specify that `Route` columns 8-11 should be character for consistency with 1-7.

```{r}
trans_ent = 
  read_csv(
    "./NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    col_types = cols(Route8 = "c", Route9 = "c", Route10 = "c", Route11 = "c")) %>% 
  janitor::clean_names() %>% 
  select(
    line, station_name, station_latitude, station_longitude, 
    starts_with("route"), entry, exit_only, vending, entrance_type, 
    ada) %>% 
  mutate(entry = ifelse(entry == "YES", TRUE, FALSE))
```

As it stands, these data are not "tidy": route number should be a variable, as should route. That is, to obtain a tidy dataset we would need to convert `route` variables from wide to long format. This will be useful when focusing on specific routes, but may not be necessary when considering questions that focus on station-level variables. 

The following code chunk selects station name and line, and then uses `distinct()` to obtain all unique combinations. As a result, the number of rows in this dataset is the number of unique stations.

```{r}
trans_ent %>% 
  select(station_name, line) %>% 
  distinct
```

The next code chunk is similar, but filters according to ADA compliance as an initial step. This produces a dataframe in which the number of rows is the number of ADA compliant stations. 

```{r}
trans_ent %>% 
  filter(ada == TRUE) %>% 
  select(station_name, line) %>% 
  distinct
```

To compute the proportion of station entrances / exits without vending allow entrance, we first exclude station entrances that do not allow vending. Then, we focus on the `entry` variable -- this logical, so taking the mean will produce the desired proportion (recall that R will coerce logical to numeric in cases like this).

```{r}
trans_ent %>% 
  filter(vending == "NO") %>% 
  pull(entry) %>% 
  mean
```

Lastly, we write a code chunk to identify stations that serve the A train, and to assess how many of these are ADA compliant. As a first step, we tidy the data as alluded to previously; that is, we convert `route` from wide to long format. After this step, we can use tools from previous parts of the question (filtering to focus on the A train, and on ADA compliance; selecting and using `distinct` to obtain dataframes with the required stations in rows).

```{r}
trans_ent %>% 
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route") %>% 
  filter(route == "A") %>% 
  select(station_name, line) %>% 
  distinct

trans_ent %>% 
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route") %>% 
  filter(route == "A", ada == TRUE) %>% 
  select(station_name, line) %>% 
  distinct
```

# Problem 2
#### read in data
```{r}
mrtw <- read_excel("./Trash Wheel Collection Data.xlsx", skip = 1, 
                   sheet = "Mr. Trash Wheel", 
                   range = cellranger::cell_cols("A:N")) %>%
                   janitor::clean_names()
```

- clean data
```{r}
tidy_mrtw <- mrtw %>%
  mutate(sports_balls = as.integer(round(sports_balls))) %>%
  mutate(type = "Mr.Trash Wheel") %>%
  mutate(year = as.integer(year), dumpster = as.integer(dumpster)) %>%
  select(everything(), weight = weight_tons,volume = volume_cubic_yards) %>%
  drop_na(dumpster)
tidy_mrtw 
```

- use similar process to read and clean professor Trash Wheel
#### read in professor data
```{r}
prtw <- read_excel("./Trash Wheel Collection Data.xlsx", skip = 1, 
                   sheet = "Professor Trash Wheel", 
                   range = cellranger::cell_cols("A:M")) %>%  
                   janitor::clean_names()
```

- clean data
```{r}
tidy_prtw <- prtw %>%
  mutate(type = "Professor Trash Wheel") %>%
  mutate(year = as.integer(year), dumpster = as.integer(dumpster)) %>%
  select(everything(), weight = weight_tons,volume = volume_cubic_yards) %>%
  drop_na(dumpster)
tidy_prtw 
```
Since there is no sports_ball variable in professor trash wheel sheet, we need to add a column for further combination of table
- Combine two dataset

```{r}
tidy_prtw$sports_balls = NA
merged_tw = rbind(tidy_mrtw,tidy_prtw)
head(merged_tw)
```

The combined dataset contains `r nrow(merged_tw)` rows and `r ncol(merged_tw)` columns. There are in total `r nrow(merged_tw)` observations in the combined dataset. There is a new column named `type` variable specifies the vessel that the observation belongs to. The key variables include: dumpster number, month, year, date, weight, volume, different kinds of trash and number, and homes powered by trash-made electricity. 

Mr. trash wheel contains ```r nrow(mrtw)``` number of data and Professor trash wheel contains ```r nrow(prtw)``` number of data. 
There are also several key values such as the weight of dumpster, which have a total weight og total of ```r sum(merged_tw$weight)``` tons. For the data collected by Mr Trash wheel, the total weight is ```r sum(filter(merged_tw,type == "Mr.Trash Wheel")$weight)``` tons, and Prof Trash wheel has a sum of ```r sum(filter(merged_tw,type == "Professor Trash Wheel")$weight)``` tons. The total volume of dumpsters is ```r sum(merged_tw$volume)``` cubic yards. The part collected by Mr Trash wheel, the total volume is ```r sum(filter(merged_tw,type == "Mr.Trash Wheel")$volume)``` cubic yards whereas Prof Trash wheel has a total of ```r sum(filter(merged_tw,type == "Professor Trash Wheel")$volume)``` cubic yards. 

The total weight of trash collected by Professor Trash Wheel is 190.12.

```{r}
merged_tw %>%
  filter(type == "Professor Trash Wheel") %>%
  select(weight) %>%
  sum()
```

The total number of sports balls collected by Mr. Trash Wheel in 2020 is 856
```{r}
merged_tw %>%
  filter(year == 2020, type == "Mr.Trash Wheel") %>%
  select(sports_balls) %>%
  sum()
```

# Problem 3
## Step 1
1. read in data
```{r}
pols_month <- read_csv(file = "./fivethirtyeight_datasets/pols-month.csv") %>%
  janitor::clean_names()
```
2. clean data 
```{r}
pols_month %>% filter(prez_gop != 0 & prez_gop != 1)
```

```{r}
pols_month %>% filter(prez_dem != 0 & prez_dem != 1)
```

There is a period of time when the boolean > 1. After checking the history we assume the president during this period is publican president.

```{r}
tidy_pol = pols_month %>% 
  separate(mon, into = c('year', 'month', 'date'), sep = '-') %>%
  mutate(month = month.name[as.numeric(month)], year = as.integer(year)) %>% 
  mutate(president = ifelse(prez_gop >= 1, "Republican", 'Democract')) %>% 
  mutate(month = factor(month, levels =  c('January','February','March','April','May','June','July',                      'August','September','October','November','December' ))) %>%
select(-prez_gop, -prez_dem, -date) %>%
arrange(year, month)
```

## Step 2
1. read in data
```{r}
snp <- read_csv(file = "./fivethirtyeight_datasets/snp.csv") %>%
  janitor::clean_names()
```

2. clean data 
```{r}
tidy_snp = snp %>% 
  separate(date, into = c('month','date','year'), sep = '/') %>% 
  mutate(month = as.integer(month), date = as.integer(date), year = as.integer(year)) %>% 
  mutate(month = month.name[as.numeric(month)]) %>% 
  mutate(year = ifelse(year <= 15, year+2000, year+1900)) %>% 
  mutate(month = factor(month, levels = c('January','February','March','April','May','June','July','August','September','October','November','December' ))) %>%
   select(year, month, close) %>%
arrange(year, month)
head(tidy_snp)
```

## Step 3
1. read in data
```{r}
unemployment <- read_csv(file = "./fivethirtyeight_datasets/unemployment.csv")%>%
  janitor::clean_names()
```

2. clean data 
```{r}
tidy_unemp = unemployment %>% 
  pivot_longer(2:13,
               names_to = "month",
               values_to = "unemployment_rate") %>% 
  mutate(month = recode(month,
                        'jan' = 'January',
                        'feb' = 'February', 
                        'mar' = 'March', 
                        'apr' = 'April', 
                        'may' = 'May', 
                        'jun' = 'June', 
                        'jul' = 'July', 
                        'aug' = 'August',
                        'sep' = 'September', 
                        'oct' = 'October', 
                        'nov' = 'November', 
                        'dec' = 'December'), 
         year = as.integer(year))  %>% 
mutate(month = factor(month, levels = c('January','February','March','April','May','June','July','August','September','October','November','December' ))) %>%
arrange(year, month)
```

## Step 4 
- combine data
```{r}
<<<<<<< HEAD
merge_df = left_join(tidy_pol, tidy_snp) %>% left_join(tidy_unemp)
=======
merge_df = left_join(tidy_pol, tidy_snp, by = c("year", "month"))
final_merge_df = left_join(merge_df, tidy_unemp, by = c("year", "month"))
>>>>>>> 19cb152f802c447dc4cc6cbc4577cca0de913059
head(merge_df)
```

The `tidy_pol` dataset includes `r count(tidy_pol)` observations and `r ncol(tidy_pol)` variables, the following variables: `r names(tidy_pol)`. 

The `tidy_snp` dataset includes `r count(tidy_snp)` observations and `r ncol(tidy_snp)` variables, the following variables: `r names(tidy_snp)`.

Tthe `tidy_unemp` dataset includes `r count(tidy_unemp)` observations and `r ncol(tidy_unemp)` variables, the following variables: `r names(tidy_unemp)`.

Combining those three datasets, we have a combined dataset that contains `r nrow(merge_df)` rows and `r ncol(merge_df)` columns.

The range of years is between `r min(merge_df$year)` and `r max(merge_df$year)`.

There are key variables in the dataset like `year`, `month`, `president`, `unemployment_rate`, and `close`. 

