---
title: "p8105_hw2_jy3270"
author: "Jingya Yu"
date: "2022-10-04"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# load all libraries
library(tidyverse)
library(readr)
library(readxl)
library(tidyr)
```

# Problem 1
#### read in data
```{r}
original <- read_csv(file = "./NYC_Transit_Subway_Entrance_And_Exit_Data.csv") %>% janitor::clean_names()
```
- clean data

```{r}
nyc_transit <- original %>%
  select(line, station_name, station_latitude:entry, vending, ada) %>%
  mutate(entry = recode(entry, "YES" = TRUE, "NO" = FALSE))
head(nyc_transit)
```

The original dataset contains `r nrow(original)` rows and `r ncol(original)` columns, namely: Division,	Line,	Station Name,	Station Latitude,	Station Longitude,	Route1,	Route2,	Route3,	Route4,	Route5,	Route6,	Route7,	Route8,	Route9,	Route10,	Route11,	Entrance Type,	Entry, Exit Only, Vending,	Staffing,	Staff Hours,	ADA,	ADA Notes, Free Crossover,	North South Street,	East West Street,	Corner,	Entrance Latitude,	Entrance Longitude,	Station Location and Entrance Location. Out of these variables, I retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance.

The cleaned dataset contains `r nrow(nyc_transit)` rows and `r ncol(nyc_transit)` columns. For data cleaning, I first clean the names of data using janitor package, then only select the required variables, and change the entry variable from character to logical variables. The data is not very clean because the route variables can be further reorganized. At the same time, some ofthe variable names contain double and have not been changed yet.

- number of distinct station

```{r}
nyc_transit %>%
  distinct(line, station_name) %>%
  nrow()
```

There are 465 distinct stations.

- number of ADA compliant stations

```{r}
nyc_transit %>%
  filter(ada == TRUE) %>%
  distinct(line, station_name) %>%
  nrow()
```

There are 84 ADA compliant stations.

- proportion of station entrances / exits without vending 

```{r}
nyc_transit %>%
  filter(vending == "NO") %>%
  pull(entry)%>%
  mean()
```

There are 37.7% of the stations entrances / exits without vending.

- number of stations serving A
```{r}
# first change all the route names to character
refored_nyc_transit = mutate(nyc_transit,across(route1:route11, as.character))

refored_nyc_transit %>%
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route") %>% 
  filter(route == "A") %>% 
  select(station_name, line) %>% 
  distinct %>% 
  nrow()
```
There are 60 stations serving A

- A train with ADA compliant
```{r}
refored_nyc_transit %>% 
  pivot_longer(route1:route11, names_to = "route_num", values_to = "route") %>%
  filter(route == "A", ada == TRUE) %>% 
  select(station_name, line) %>% 
  distinct() %>%
  nrow()
```

There are 17 stations serving A with ADA compliant.

# Problem 2
#### read in data
```{r}
mrtw <- read_excel("./Trash Wheel Collection Data.xlsx", skip = 1, 
                   sheet = "Mr. Trash Wheel", 
                   range = cellranger::cell_cols("A:N")) %>%
                   janitor::clean_names()
```

- clean data
```{r}
tidy_mrtw <- mrtw %>%
  mutate(sports_balls = as.integer(sports_balls)) %>%
  mutate(type = "Mr.Trash Wheel") %>%
  mutate(year = as.integer(year), dumpster = as.integer(dumpster)) %>%
  select(everything(), weight = weight_tons,volume = volume_cubic_yards) %>%
  drop_na(dumpster)
tidy_mrtw 
```

- use similar process to read and clean professor Trash Wheel
#### read in professor data
```{r}
prtw <- read_excel("./Trash Wheel Collection Data.xlsx", skip = 1, 
                   sheet = "Professor Trash Wheel", 
                   range = cellranger::cell_cols("A:M")) %>%  
                   janitor::clean_names()
```

- clean data
```{r}
tidy_prtw <- prtw %>%
  mutate(type = "Professor Trash Wheel") %>%
  mutate(year = as.integer(year), dumpster = as.integer(dumpster)) %>%
  select(everything(), weight = weight_tons,volume = volume_cubic_yards) %>%
  drop_na(dumpster)
tidy_prtw 
```
Since there is no sports_ball variable in professor trash wheel sheet, we need to add a column for further combination of table
- Combine two dataset

```{r}
tidy_prtw$sports_balls = NA
merged_tw = rbind(tidy_mrtw,tidy_prtw)
head(merged_tw)
```

The combined dataset contains `r nrow(merged_tw)` rows and `r ncol(merged_tw)` columns. There are in total `r nrow(merged_tw)` observations in the combined dataset. There is a new column named `type` variable specifies the vessel that the observation belongs to. The key variables include: dumpster number, month, year, date, weight, volume, different kinds of trash and number, and homes powered by trash-made electricity. 

Mr. trash wheel contains ```r nrow(mrtw)``` number of data and Professor trash wheel contains ```r nrow(prtw)``` number of data. 

There are also several key values such as the weight of dumpster, which have a total weight og total of ```r sum(merged_tw$weight)``` tons. For the data collected by Mr Trash wheel, the total weight is ```r sum(filter(merged_tw,type == "Mr.Trash Wheel")$weight)``` tons, and Prof Trash wheel has a sum of ```r sum(filter(merged_tw,type == "Professor Trash Wheel")$weight)``` tons. The total volume of dumpsters is ```r sum(merged_tw$volume)``` cubic yards. The part collected by Mr Trash wheel, the total volume is ```r sum(filter(merged_tw,type == "Mr.Trash Wheel")$volume)``` cubic yards whereas Prof Trash wheel has a total of ```r sum(filter(merged_tw,type == "Professor Trash Wheel")$volume)``` cubic yards. 

The total weight of trash collected by Professor Trash Wheel is 190.12
```{r}
merged_tw %>%
  filter(type == "Professor Trash Wheel") %>%
  select(weight) %>%
  sum()
```

The total number of sports balls collected by Mr. Trash Wheel in 2020 is 856
```{r}
merged_tw %>%
  filter(year == 2020, type == "Mr.Trash Wheel") %>%
  select(sports_balls) %>%
  sum()
```

